---
title: "ECN Taxa List"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{taxa}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  echo = FALSE
)
library(macroinvertebrateMetrics)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(magrittr)
library(knitr)
```


```{r}
# format taxa table types
taxa <- macroinvertebrateTaxa
taxa$TAXON_NAME <- as.character(taxa$TAXON_NAME)
taxa$TL2_5_TAXON <- as.character(taxa$TL5_TAXON)
```

# Intro

What list should we use for ECN sites?

# ECN Vs TL5

What's in TL5 but not in TL_ECN

- Mainly Beetle and Corixidae species

```{r}
tl5 <- filter(taxa, TAXON_NAME == TL5_TAXON)
tl5_not_in_ecn <- filter(tl5, ECN_TL == FALSE)
tl5_not_in_ecn$TAXON_NAME

```

What's in ECN and not in TL5

- Mainly Oligochaeta families and Beetle families not found in TL5

```{r}

ecn <- filter(taxa, ECN_TL == TRUE)
ecn_not_in_tl5 <- filter(ecn, !TAXON_NAME %in% tl5$TAXON_NAME)
ecn_not_in_tl5$TAXON_NAME
```

## TL4

```{r}
tl4 <- filter(taxa, TAXON_NAME == TL4_TAXON)
ecn <- filter(taxa, ECN_TL == TRUE)
```

TL4 is double the length of ECN list (`r length(tl4$TAXON_NAME)` Vs `r length(ecn$TAXON_NAME)`). TL4 is much more resource intensive than ECN list. Don't recommend using TL4. 

## So what list should we use?

What we need is TL5 + ECN? For a ECN analysis which can be 'downgraded' to TL5 and used for AWIC/LAMM or other TL5 level metrics. 

However, we need to take away family level options where genus/species options are available in ECN/TL5, for instance:

- Remove Oligochaeta (TL5) as option and replace with sub families from ECN list (Naididae, Lumbriculidae etc)
- Add *Corixa affinis* (TL5) and remove Corixidae (ECN) 

ECN familes to keep:
```{r}
ecn_family <- filter(ecn, RANK == "Family") 
tl5_genus_species <- filter(tl5, FAMILY %in% ecn$TAXON_NAME)
ecn_family_keep <- filter(ecn_family, !TAXON_NAME %in% tl5_genus_species$FAMILY)
ecn_family_keep
```

TL5 families to keep
```{r}
tl5_family <- filter(tl5, RANK == "Family") 
ecn_genus_species <- filter(ecn, FAMILY %in% ecn$TAXON_NAME)
tl4_family_keep <- filter(tl5_family, !TAXON_NAME %in% ecn_genus_species$FAMILY)
tl4_family_keep
```

# Not Active

Check column `active_data_entry` = FALSE
  
ECN taxa not active:
```{r}
filter(ecn, ACTIVE_DATA_ENTRY == FALSE) %>%  select(TAXON_NAME) %>% kable(row.names = FALSE)
```

TL5 taxa not active:
```{r}
filter(tl5, ACTIVE_DATA_ENTRY == FALSE) %>%  select(TAXON_NAME) %>% kable(row.names = FALSE)
```







